---
title: "Utilizing ChatGPT for Social Psychology Research"
author: "Bryan West"
institute: "Psychology Methods Group"
date: "April 24, 2023"
format: 
  revealjs:
    theme: night
    center: true
    slide-number: true
    show-notes: false
    smaller: true
    slide-level: 4
    toc-depth: 2
    css: custom.css
    link-external-newwindow: true
---

## Introduction

::: incremental
-   What is ChatGPT?
-   Benefits and use cases for social psychology researchers
-   Examples and demonstrations
:::

::: notes
-   Briefly introduce ChatGPT and its capabilities
-   Explain the benefits it can provide to social psychology graduate students
-   Mention that the presentation will walk through specific examples and
    demonstrations
:::

### What is ChatGPT?

::: incremental
-   An AI-powered language model by OpenAI
-   Capable of sophisticated language processing
-   Can understand and generate human-like text
:::

### Capabilities of ChatGPT

::: incremental
-   Context understanding and relevant information
    -   Processes user inputs to provide contextually appropriate responses
    -   Generates detailed and accurate information
-   Research and writing assistance
    -   Helps with various research-related tasks, from literature reviews to
        manuscript editing
-   Creative idea generation
    -   Suggests innovative ideas for research questions and methods
-   Tailored assistance through learning
    -   Adapts to user inputs for more personalized assistance over time

:::

### Caveats

::: incremental
-   ChatGPT is still in its early stages. Lots of things will probably change in the near future. 
-   Some of the features I will be demonstrating are only available to alpha testers, but they will likely become available to the general public soon.
-   GPT-4 is not yet available on OpenAI's free account tier, but GPT-3.5 is still very powerful.
    -   Many of the examples I will be showing you today were created with GPT-4.
-   This entire presentation was created with **substantial** assistance from GPT-4.

:::

### Use Cases

::: incremental
-   Coding assistant
    -   Write code to accomplish a specified task
    -   Explicate and comment existing code
    -   Refactoring
    -   Troubleshooting assistance
-   Accelerating the research process
    -   Assist with literature reviews
    -   Identify gaps in literature
    -   Summarizing key findings
-   Assisting with research question development
    -   Explore under-researched areas and suggest new perspectives
    
:::
    
### Use Cases

::: incremental
-   Guiding research design
    -   Recommendations for methdologies, sample sizes, and ethical
        considerations
    -   Design survey questions and interview scripts/guides
-   Supporting data analysis
    -   Suggest appropriate statistical tests
    -   Assistance in interpreting results
-   Aiding in the interpretation of findings
    -   Identify trends and discuss implications for theories and interventions
-   Providing writing assistance
    -   Draft research papers
    -   Proofreading
    -   Editing for style and clarity

:::

### 1. Coding Assistance

-   Write code to accomplish a specified task
-   Explicate and comment existing code
-   Refactoring
-   Troubleshooting assistance

#### Write code to accomplish a speciic task

**Task:** Conduct a t-test to compare the means of two groups in a dataset

**R:**

```{r}
#| label: r-code-ttest
#| echo: true
#| eval: false

# Import data
data <- read.csv("your_data.csv")

# Conduct t-test
t_test_result <- t.test(data$group1, data$group2)

# Print results
print(t_test_result)
```

**Python:**

```{python}
#| label: python-code-ttest
#| echo: true
#| eval: false

import pandas as pd
from scipy.stats import ttest_ind

# Import data
data = pd.read_csv("your_data.csv")

# Conduct t-test
t_test_result = ttest_ind(data['group1'], data['group2'])

# Print results
print(t_test_result)

```

#### Explicate/comment existing code

**Task:** Add coments to explain the following code snippet.

**R:**

```{r}
#| label: r-code-comment
#| echo: true
#| eval: false

# Load required libraries
library(ggplot2)

# Create a scatterplot of variable1 against variable2, with color representing group
ggplot(data, aes(x = variable1, y = variable2, color = group)) + 
  geom_point() + 
  theme_minimal() +
  labs(title = "Scatterplot of Variable1 vs Variable2 by Group",
       x = "Variable 1",
       y = "Variable 2",
       color = "Group")
```

**Python:**

```{python}
#| label: python-code-comment
#| echo: true
#| eval: false

import seaborn as sns
import matplotlib.pyplot as plt

# Create a scatterplot of variable1 against variable2, with color representing group
scatter_plot = sns.scatterplot(data=data, x='variable1', y='variable2', hue='group')

# Customize plot appearance
scatter_plot.set_title('Scatterplot of Variable1 vs Variable2 by Group')
plt.xlabel('Variable 1')
plt.ylabel('Variable 2')
plt.legend(title='Group')
plt.show()
```

#### Refactor code

Task: Refactor the following code snippet to use a function.

**R:**

```{r}
#| label: r-code-refactor
#| echo: true
#| eval: false

# Original code
data$z_score_group1 <- (data$group1 - mean(data$group1)) / sd(data$group1)
data$z_score_group2 <- (data$group2 - mean(data$group2)) / sd(data$group2)

# Refactored code using a function
z_score <- function(x) {
  return((x - mean(x)) / sd(x))
}

data$z_score_group1 <- z_score(data$group1)
data$z_score_group2 <- z_score(data$group2)
```

**Python:**

```{python}
#| label: python-code-refactor
#| echo: true
#| eval: false

# Original code
data['z_score_group1'] = (data['group1'] - data['group1'].mean()) / data['group1'].std()
data['z_score_group2'] = (data['group2'] - data['group2'].mean()) / data['group2'].std()

# Refactored code using a function
def z_score(x):
    return (x - x.mean()) / x.std()

data['z_score_group1'] = z_score(data['group1'])
data['z_score_group2'] = z_score(data['group2'])
```

#### Troubleshoot code

**Task:** Identify and fix issues in the following code snippet.

**R:**

```{r}
#| label: r-code-troubleshoot
#| echo: true
#| eval: false

# Original code with an issue
data <- read.csv("your_data.csv")
mean_group1 <- mean(data$group1, na.rm = Ture) # Typo in 'True'

# Fixed code
data <- read.csv("your_data.csv")
mean_group1 <- mean(data$group1, na.rm = TRUE)
```

**Python:**

```{python}
#| label: python-code-troubleshoot
#| echo: true
#| eval: false

# Original code with an issue
import pandas as pd
data = pd.read_csv("your_data.csv")
mean_group1 = data['group1'].mean(na.rm=True)  # Incorrect parameter for .mean()

# Fixed code
import pandas as pd
data = pd.read_csv("your_data.csv")
mean_group1 = data['group1'].mean(skipna=True)
```

#### Example: Zotero Assistant

I asked ChatGPT to help me write a script that would summarize article
abstracts, selecting the articles by their folder location in my Zotero library.

-   I did not know how to code in Python.
-   I did not know how to use OpenAI's API interface.
-   This lack of knowledge did not matter.

### 2. Literature Review

::: incremental
-   Search for literature relevant to research question
-   Summarizing key findings
-   Analyze gaps in research literature
-   Example prompt:
    -   *"Find relevant research literature in social psychology on the impact
        of social media on self-esteem, providing a summary of key findings and
        identifying research gaps."*
    -   [ChatGPT's response](https://chat.openai.com/c/2555d9a6-7a61-49b1-a199-d25f09115b13){target="_blank"}
-   GPT-3.5 will frequently "hallucinate" sources that are not real. 
-   GPT-4 does this much less frequently, but it is still prudent to check any sources it references to make sure that they are correct and relevant.
    -   GPT-4 was trained on data through September 2021, so any sources published after then will not be accessible to it.

:::

::: notes
-   Describe the process of using ChatGPT to find relevant literature
-   Explain how ChatGPT can summarize key findings from studies
-   Note that GPT-4 only has data through September 2021, but that future models
    will be more up-to-date or even be capable of searching the web.
:::

#### How does ChatGPT identify gaps in existing literature?

::: incremental
-   I asked it this exact question.
:::

#### How does ChatGPT identify gaps in existing literature?

ChatGPT identifies gaps in existing literature by analyzing the content of
available research articles, reviews, and other scholarly sources. It processes
the text to understand the main findings, methodologies, and themes discussed in
the literature. By recognizing patterns and trends within this information,
ChatGPT can identify areas that have not been extensively explored or addressed
in the existing body of work.

<!-- ### How does ChatGPT identify research gaps? -->

To identify gaps, ChatGPT may consider:

1.  Unexplored research questions or areas: ChatGPT can detect topics or
    questions that have not been addressed or only briefly mentioned in the
    existing literature.

2.  Limitations of previous studies: By reviewing the limitations sections in
    research papers, ChatGPT can pinpoint methodological issues, small sample
    sizes, or other shortcomings that may warrant further investigation.

3.  Contradictory findings: ChatGPT can identify inconsistencies or
    disagreements between different studies, which could suggest a need for more
    research to clarify these inconsistencies.

4.  Underrepresented populations or contexts: ChatGPT can detect when certain
    populations, settings, or contexts have not been adequately represented in
    the literature, indicating an opportunity to expand the knowledge base.

5.  Calls for future research: By reviewing the conclusions or future research
    sections of papers, ChatGPT can identify areas where authors have explicitly
    suggested the need for further investigation.

It is important to note that ChatGPT's ability to identify gaps in the
literature depends on the quality and quantity of the sources it has been
trained on, as well as the clarity and specificity of user inputs. While it can
provide valuable insights, it is still recommended that researchers verify
ChatGPT's findings by consulting primary sources and conducting their own
critical analysis of the literature.

### 3. Research Question Development

::: incremental
-   Generate research question based on identified gaps in literature
-   Example: "What factors buffer against the negative effects of social media
    on self-esteem?"
:::

::: notes
-   Explain how ChatGPT can identify gaps in existing literature
-   Describe how it can suggest new perspectives and research questions
-   Discuss the example research question on online social support networks and
    stress managementp
:::

### 4. Research Design and Study Materials

::: incremental
-   Generate survey questions, stimulus materials, vignettes, etc.
-   Evaluate survey questions and suggest improvements.
-   Generate possible study designs, given a research question.
-   Example prompt: "Suggest some research designs to identify factors that
    buffer against the negative effects of social media on self-esteem that
    could help inform interventions and promote more positive online
    experiences. Be specific about methodologies, determine appropriate criteria
    and sample sizes, and explain your reasoning and analysis process."
:::

::: notes
-   Explain how ChatGPT can provide guidance on research methodologies
-   Describe its ability to help with sample size determination and selection
    criteria
:::

#### Investigating Factors Buffering Against Negative Effects of Social Media on Self-Esteem: Research Designs

##### 1. Longitudinal Survey Design

Methodology: Participants will be recruited from a diverse range of age groups,
gender identities, and cultural backgrounds. Researchers can use online
platforms or a combination of online and offline methods to recruit a large,
diverse sample (e.g., at least 1,000 participants). Participants will be asked
to complete an initial survey measuring their self-esteem, social media use, and
potential buffering factors, such as social support, digital literacy, and
coping strategies. Follow-up surveys will be conducted at 6-month intervals for
at least two years.

Analysis: Researchers will use linear mixed-effects modeling or other
appropriate longitudinal analysis methods to examine the relationship between
self-esteem, social media use, and potential buffering factors over time.
Moderation analyses will be performed to determine whether these factors
mitigate the negative effects of social media on self-esteem.

#### Investigating Factors Buffering Against Negative Effects of Social Media on Self-Esteem: Research Designs

##### 2. Experimental Design with Randomized Controlled Trial:

Methodology: Participants will be randomly assigned to either an intervention
group or a control group. The intervention could be an online program designed
to teach digital literacy, promote healthy social media use, or develop coping
strategies for dealing with negative online experiences. Sample size should be
determined using power analysis to ensure adequate statistical power for
detecting group differences (e.g., at least 200 participants per group).

Criteria: Researchers should develop a standardized set of criteria for
selecting participants, such as age range, social media usage patterns, and
baseline self-esteem levels. Additionally, eligibility criteria might include
the absence of any pre-existing mental health conditions that could confound the
study results.

Analysis: Researchers will use analysis of covariance (ANCOVA) or other
appropriate statistical methods to compare self-esteem levels between the
intervention and control groups, both immediately after the intervention and at
follow-up assessments (e.g., 3 months, 6 months, and 12 months
post-intervention). Mediation analyses can be performed to explore whether the
intervention's effects on self-esteem are mediated by changes in the potential
buffering factors.

#### Investigating Factors Buffering Against Negative Effects of Social Media on Self-Esteem: Research Designs

##### 3. Qualitative Research Design:

Methodology: Researchers can conduct semi-structured interviews or focus groups
with a diverse sample of social media users (e.g., 30-50 participants) to
explore their experiences, perceptions, and coping strategies related to social
media use and self-esteem.

Criteria: Participants should represent a diverse range of age groups, gender
identities, and cultural backgrounds. Researchers can use purposive sampling to
recruit participants who have varying levels of social media usage and
self-esteem.

Analysis: Researchers will use thematic analysis or other appropriate
qualitative analysis methods to identify common themes, patterns, and factors
that buffer against the negative effects of social media on self-esteem. This
qualitative approach can provide valuable insights into the participants'
experiences and help generate hypotheses for future quantitative research.

#### Summarizing Content

That was a lot of text. Let's ask ChatGPT to summarize and convert it to a
format that works for a presentation/slideshow.

Prompt: "Summarize these designs into a format that could be presented in a
slideshow."

#### Overview

-   Longitudinal Survey
-   Experimental Design with Randomized Controlled Trial
-   Qualitative Research Design

#### Longitudinal Survey

-   Objective: Examine the relationship between self-esteem, social media use,
    and buffering factors over time
-   Large, diverse sample (e.g., 1,000 participants)
-   Surveys at baseline and 6-month intervals for at least two years
-   Data analysis: Linear mixed-effects modeling, moderation analyses

#### Experimental Design with Randomized Controlled Trial

-   Objective: Evaluate the effectiveness of an intervention to buffer against
    negative effects of social media on self-esteem
-   Intervention group vs. control group
-   Sample size determined by power analysis (e.g., at least 200 participants
    per group)
-   Data analysis: ANCOVA, mediation analyses

#### Qualitative Research Design

-   Objective: Explore experiences, perceptions, and coping strategies related
    to social media use and self-esteem
-   Semi-structured interviews or focus groups
-   Diverse sample of social media users (e.g., 30-50 participants)
-   Data analysis: Thematic analysis or other qualitative analysis methods

#### Conclusion

-   These research designs can help identify factors that buffer against the
    negative effects of social media on self-esteem
-   Findings can inform interventions and strategies for promoting positive
    online experiences

### 5. Data Analysis

::: incremental
-   Data cleaning and preprocessing
    -   You can copy/paste CSV-formatted text into ChatGPT and have it perform
        analyses, clean responses, etc.
-   Check that data meet assumptions required for specific analysis methods
    -   Normality, homeoscedasticity, and linearity for regression analyses
    -   If assumptions are not met, suggest data transformations or alternative
        analysis methods
:::

::: notes
-   Explain how ChatGPT can help with data cleaning and preprocessing
-   Describe its ability to recommend appropriate statistical tests based on
    research questions and data
:::

### 6. Interpretation of Findings

::: incremental
-   Identifying trends and patterns
-   Implications for social psychology
-   Example:
    -   Interpreting negative correlation between social media usage and
        self-esteem
:::

::: notes
-   Explain how ChatGPT can help interpret results and identify trends,
    patterns, and relationships
-   Describe its ability to discuss implications of findings for social
    psychology theory and practice
-   Discuss the example interpretation of the negative correlation between
    social media usage and self-esteem
:::

#### Interpretation of Findings: Negative correlation between social media useage and self-esteem

Prompt: *"Assume that I ran a study that found a negative correlation between
social media usage and self-esteem. Discuss some possible implications for
social psychology that would be of note and important to discuss in a research
article about this study."*

If your study found a negative correlation between social media usage and
self-esteem, it would have several important implications for the field of
social psychology. In a research article discussing the study, it would be vital
to address the following points:

1.  The role of social comparison: A negative correlation between social media
    usage and self-esteem might be partly explained by social comparison
    processes. Social media platforms often showcase idealized images and
    achievements of others, which can lead to upward social comparisons and
    feelings of inadequacy (Festinger, 1954). Highlighting the importance of
    social comparison in the context of social media can help researchers better
    understand its role in self-esteem.

2.  Impact on well-being: The relationship between social media use and
    self-esteem has implications for overall well-being, as self-esteem is
    closely linked to mental health and life satisfaction (Orth & Robins, 2013).
    Discussing the potential consequences of reduced self-esteem due to social
    media use, such as increased risk for depression, anxiety, and social
    isolation, can underscore the broader impact of these findings.

3.  Vulnerable populations: Certain populations, such as adolescents and young
    adults, might be more susceptible to the negative effects of social media on
    self-esteem due to developmental factors (e.g., identity formation) and
    increased social media use. Identifying vulnerable populations and
    discussing the specific risks they face can help inform targeted
    interventions and support.

#### Interpretation of Findings: Negative correlation between social media useage and self-esteem

4.  Individual differences: The negative correlation between social media usage
    and self-esteem might not be universal. It's essential to explore individual
    differences, such as personality traits, coping strategies, or social
    support, that may moderate the relationship and help identify factors that
    buffer against these negative effects.

5.  Quality of social media use: The nature of social media use, rather than its
    mere quantity, might be a crucial factor in the relationship with
    self-esteem. Discussing the role of active versus passive use, engagement in
    supportive communities, and exposure to diverse content can help researchers
    refine their understanding of how different types of social media
    experiences affect self-esteem.

6.  The need for interventions: The observed negative correlation highlights the
    importance of developing interventions to mitigate the detrimental effects
    of social media on self-esteem. Discussing potential strategies, such as
    promoting digital literacy, encouraging healthy social media habits, and
    enhancing online social support, can contribute to the development of
    effective interventions.

7.  Future research directions: Lastly, discussing the limitations of your study
    and suggesting future research directions can help advance the understanding
    of the relationship between social media use and self-esteem. This might
    include exploring causal relationships through experimental designs,
    identifying protective factors, or examining the impact of specific social
    media platforms and features.

By discussing these implications in a research article, you can contribute to
the ongoing conversation about the role of social media in shaping self-esteem
and inform the development of interventions and guidelines that promote
healthier online experiences.

### 7. Writing Assistance

::: incremental
-   Drafting and editing research papers
-   Ensuring clear and concise writing
-   Example: Introduction for a research article about the relationship between social media use and self-esteem
    -   Prompt: *"Write an outline for the introduction section of a social psychology research article investigating the relationship between social media use and self-esteem, and what factors help buffer against the negative effects of social media on self esteem. After you have written the outline, write a draft of the introduction section. The style should be appropriate for a professional academic research journal in the field of social psychology. Make sure to include references (in APA style) to empirical research articles that are relevant to the subject."*
    -   [ChatGPT's response](https://chat.openai.com/c/2555d9a6-7a61-49b1-a199-d25f09115b13)

:::

::: notes
-   Explain how ChatGPT can provide support in drafting and editing research
    papers, conference presentations, and thesis chapters
-   Describe its ability to ensure clear, concise, and well-organized writing
-   Discuss the example of drafting an introduction for a paper on the
    relationship between social media use and self-esteem

:::

## Conclusion

ChatGPT can be a powerful resource for you as a researcher.

::: incremental
-   Coding support
-   Research and literature review assistance
-   Idea generation
-   Study design and survey material creation
-   Writing assistance (outlining, editing, proofreading, style)
:::
